Hierarchical recurrent models reveal latent states of neural activity in \textit{C. elegans}

### 2018-01-10 drafts/elife/draft.tex
syncing

\begin{abstract}
  Recent advances in neural recording technologies have enabled
  simultaneous measurements of the majority of head ganglia neurons in
  both immobilized and freely-behaving
  \celegans~\citep{schrodel2013brain, prevedel2014simultaneous,
    nguyen2016whole}.  The dynamics of neural activity shed light on
  how \celegans~processes sensory information and generates motor
  activity.  To understand these dynamics, we develop recurrent state
  space models that decompose complex time-series into segments with
  simple, linear dynamics. We incorporate these models into a robust,
  hierarchical framework for combining information across whole-brain
  recordings of many worms.  Using this framework, we reveal latent
  states of population neural activity, along with the discrete
  behavioral modes that drive dynamics in this latent state space.  We
  find stochastic transition patterns between these discrete
  behavioral modes, and we see that transition probabilities are
  determined by a combination of current brain state and environmental
  cues.  In addition to quantifying neural dynamics, this
  probabilistic framework aids in neural identification---currently a
  laborious, manual task---and reveals clusters of neurons that are
  similarly tuned in latent state space.  Finally, we find a
  significant overlap between our inferred states and the
  manually-labeled states of~\citet{kato2015global}
  and~\citet{nichols2017global}, which were shown to correspond to
  different behaviors, like forward crawling, reversals, and
  turns. Our methods automatically discover and quantify these
  behaviorally-meaningful states directly from neural activity,
  yielding powerful new tools for neuro-behavioral analysis.
\end{abstract}

@@

@@

We seek a parsimonious characterization of the dynamics of neural
activity in the head ganglia of \celegans. Recent advances in optical
imaging offer simultaneous recordings of most of the head ganglia
neurons, but we must surmount three main challenges in order to learn
a dynamics model from these data.  First, we need to design a model
flexible enough to capture nonlinear dynamics yet interpretable enough
to reveal simplifying structure when it exists.  Moreover, any model
will necessarily be an approximation to the true neural dynamics, and
as such we need to be robust to data that does not quite match model
predictions.  Finally, our model must combine partial recordings from
separate organisms to learn shared neural dynamics yet still allow for
individual variability. We construct a recurrent, robust, and
hierarchical state space model that addresses these challenges,
decomposing complex dynamics into a collection of simpler pieces that
can be learned from noisy, partial recordings with trial-to-trial
variability.

We study the data presented in~\citet{kato2015global} and
\citet{nichols2017global}, which consist of a collection of optical
recordings of calcium fluorescence in immobilized \celegans. The
recordings are 18 minutes long with a frame rate of about 3Hz.  On
average, about 100 single units are extracted from each
recording. \celegans~is genetically stereotyped and each of its 302
neurons has a unique label (e.g. \textsf{AVAL}). On average, 30 single
units could be unambiguously labeled in each
recording. Fig.~\ref{fig:model}a illustrates this type of data. The
three panels represent recordings from three separate worms, and the
rows correspond to different neuron labels.  Neural activity is
measured for only a subset of neurons in each worm.

\begin{figure}[t!]
\centering%
\includegraphics[width=6in]{figures/v3/figure1} 
\caption{
  \textit{A probabilistic framework for combining
  information across multiple worms in order to learn a shared
  dynamical system for \celegans.}
  \textbf{a.}~The data consists of observed neural activity collected
  from many different worms.  In each recording we observe a different
  subset of labeled neurons in the head ganglia.
  \textbf{b.}~We model the observed activity as a linear function of
  an underlying, continuous, and typically low-dimensional latent state,
  here shown as a two-dimensional signal for each recording.
  \textbf{c.}~The dynamics of these continuous states are governed by
  a co-evolving discrete state.
  \textbf{d.}~These discrete states index into a library of linear
  dynamics functions for each worm, here shown as maps from~$\reals^2$ to~$\reals^2$.
  Each dynamics function is selectively used in different regions of
  continuous latent space, as shown by the different intensities of
  the vector field.
  \textbf{e.}~To increase robustness to model misspecification, we
  use a multivariate-t distribution for the noise in the continuous
  state updates.
  \textbf{f.}~Each worm has its own dynamics library, but they are
  constrained to be close to the shared, canonical library.  This
  combines information to learn shared structure while also allowing
  for worm-to-worm variability.
  \textbf{g.} Likewise, we allow each worm to have its own observation
  variances to account for individual differences that may arise from,
  for example, variable levels of fluorescent protein expression.
  Missing neurons are denoted by gray boxes. 
}
\label{fig:model}
\end{figure}

@@

\begin{figure}[t!]
\centering
\includegraphics[width=6in]{figures/v3/figure2} 
\caption{ \textit{Hierarchical state space models smooth noisy
    recordings and predict the activity of unseen neurons.}  We show
  the neural activity (differenced Ca++ traces) for two of the
  five worms from the \citet{kato2015global} dataset.  Of these five
  worms, worm 1 (left) has the fewest labeled neurons and worm 5
  (right) has the most.  Green lines are the measured activity of the
  labeled neurons; black lines are the smoothed activities predicted
  by the model for the 61 neurons that were labeled in at least one
  worm.  The model uses correlations learned across all recordings to
  predict the activity of unlabeled neurons, like \textsf{SMBDL} in
  worm 1.  The neurons have been clustered based on similarity, as
  discussed in Fig.~\ref{fig:clustering}.  }
\label{fig:smoothing}
\end{figure}

@@

The degree to which the smoothed activity matches the observed
activity is governed by the dimensionality of the latent state space.
We estimate this dimensionality with cross validation, reserving the
last 20\% of the data (3.6 min) for testing and measuring the marginal
log likelihood of the held-out data as we increase the latent
dimensionality. We find that the test likelihood peaks with a
10-dimensional latent state (Fig.~\ref{fig:supp_emissions}c).

Moreover, the likelihood is substantially higher in the model that
includes per-worm observation variances compared to the model with
shared variances for all worms. Fig.~\ref{fig:supp_emissions}d shows
the inferred variance for each worm in the \citet{kato2015global}
dataset.  Neurons that were not identified are marked with a gray box.
This shows the extent of the missing data and the need for
hierarchical modeling.\todo{make supplement figure}

@@

\begin{figure}[t!]
\centering
\includegraphics[width=6in]{figures/v3/figure3B} 
\caption{ \textit{The mapping from continuous latent states to neural
    activity reveals clusters of functionally similar neurons.}
  \textbf{a.} The emission matrix is a linear map from latent states;
  the rows can be seen as ``tuning vectors'' that specify how each
  neuron responds to changes in the latent state.  These tuning
  vectors cluster into discrete groups (separated by black lines),
  indicating that there are subpopulations of similarly tuned, and
  hence highly correlated, neurons. Neurons are ordered as in
  Fig.~\ref{fig:smoothing}. \textbf{b.}~The pairwise similarity between
  tuning vectors supports this clustering and shows differences
  between subpopulations (clusters have been sorted to make this as
  close to diagonal as possible).  \textbf{c.}~The tuning similarity
  reflects correlations between neurons, but since it is learned as
  part of a hierarchical model, it predicts similarities for all pairs
  of neurons, even pairs that are never simultaneously recorded (gray,
  off-diagonal boxes).  \textbf{d.}~We show the spatial locations of
  neurons from each cluster, as specified by WormAtlas.  Neurons in
  the same cluster are not necessarily close to one another in space,
  though some clusters are spatially localized. Black bars denote
  approximate locations of the anterior (AG), dorsal (DG), ventral
  (VG) and retrovesicular ganglia.}
\label{fig:clustering}
\end{figure}

@@

\begin{figure}[t!]
\centering
\includegraphics[width=6in]{figures/v3/figure_id} 
\caption{ \textit{The predicted neural activity is a useful aid for
    neural identification.}  \textbf{a.} The model uses observed and
  labeled activity (green traces) to predict the activity of unlabeled
  neurons, as in Fig.~\ref{fig:smoothing}. \textbf{b.} Alongside the
  labeled activity, we also have traces from many neurons that could
  not be unambiguously labeled. To test whether the model's
  predictions could aid in labeling neurons, we conducted a synthetic
  experiment, withholding the activity of some labeled neurons. Here,
  the red trace is the withheld activity of VB01 and the blue traces
  are the activity of other unlabeled neurons.  Together, these make
  up a pool of candidates for the VB01 label.  \textbf{c.} For each
  candidate, we measured the correlation between its activity and the
  predicted activity of VB01.  \textbf{d.} Repeating this experiment
  50 times, we found that the correct candidates tend to be much more
  correlated with the model's predicted activity than the other,
  incorrect candidates.  Thus, the model's predictions serve as a
  useful aid for labeling neurons.  }
\label{fig:id}
\end{figure}

@@

\begin{figure}[t!]
\centering
\includegraphics[width=6in]{figures/v3/figure4} 
\caption{ \textit{The hierarchical SLDS automatically segments neural
    activity based on its dynamics.}  \textbf{a.}~The top three
  dimensions of the inferred continuous latent states for each of the
  five worms, color coded by the corresponding discrete latent state.
  \textbf{b.}~The inferred segmentations (top) align with the manual
  segmentations (bottom) of~\citet{kato2015global}, though some
  manually labeled states are split into multiple inferred states
  (e.g. forest green state is often split into green and brown) and
  others are combined (e.g. light purple, light blue, and white are
  often combined into pink).  \textbf{c.}~We quantified the overlap
  between the manual and inferred states and found a close
  correspondence.  As~\citet{kato2015global} showed, the manually
  labeled states map onto different behaviors (\textsf{REVSUS}:
  sustained reversal; \textsf{SLOW}: slow forward crawl; \textsf{VT}:
  ventral turn; \textsf{FWD}: forward crawl; \textsf{DT}: dorsal turn;
  \textsf{REV1, REV2}: reversals; \textsf{NOSTATE}:
  undetermined). The correspondence between inferred and manually
  labeled states implies that hierarchical state space models can
  infer behaviorally meaningful states directly from neural activity.}
\label{fig:syllables}
\end{figure}

@@

Finally, Fig.~\ref{fig:dynamics}c shows the eigenspectra of the linear
dynamics for each worm and each state.  Imaginary eigenvalues give
rise to rotational dynamics, and the magnitude of the eigenvalues
determines whether the state is stable or not.  If the magnitude is
greater than one (outside the unit disk denoted by the black arc),
the dynamics produce exponential growth in the latent state space; if
the eigenvalues are all less than unit magnitude, the dynamics converge
to a fixed point.  As expected, the rotational states do indeed have
more imaginary eigenvalues, and most states have eigenvalues with
magnitude near one.  Some states, like the brown state, have eigenvalues
larger than one, which produce unstable dynamics, though this is
somewhat balanced by the recurrent extensions discussed in the next
section.  Comparing across rows, we see only minor differences from worm
to worm, substantiating our hierarchical modeling assumptions.

\begin{figure} [b!]
  \caption{ \textit{(Next page.) Discrete states are characterized by
      simple linear dynamics that give rise to interpretable patterns
      of neural activity.}  \textbf{a.}~Each discrete state is
    associated with a canonical set of linear dynamics, which can be
    viewed as mappings from~$\reals^{10}$ to~$\reals^{10}$.  We show
    the mappings for the first three dimensions here, with vectors
    placed at locations where each state was used.  \textbf{b.}~We
    simulated four latent state trajectories with these linear
    dynamics from different starting locations, and then we mapped
    these latent state trajectories through the emission matrix to
    compute a time course of resulting activations for each cluster of
    neurons.  \textbf{c.}~The eigenspectra of the dynamics matrices
    provide insight into the nature of the states: imaginary eigenvalues
    give rise to rotational dynamics and the magnitude determines
    whether the stability of the dynamics. The arc denotes
    the unit magnitude line; eigenvalues to the left are contractive
    and to the right are expansive. Looking across rows, we
    see how the spectra vary from worm to worm. }
\end{figure}
\addtocounter{figure}{-1}

\begin{figure}[h!]
  \centering
  \vspace{-.5in}
  \includegraphics[width=6in]{figures/v3/figure5B}
  \caption{Caption on preceding page.}
  \label{fig:dynamics}
\end{figure}

@@

\begin{figure}[t!]
\centering
\includegraphics[width=6in]{figures/v3/figure6} 
\caption{ \textit{Discrete state transitions show a chain-like structure
    with non-Markovian durations.}  \textbf{a.}~Counts of inferred transitions
  from one state to another. The states are permuted to emphasize the
  feed-forward structure.  \textbf{b.}~An alternative view of the
  transitions; each state is a node in a graph and edges denote
  high-probability transitions.  This shows transitions between
  forward and reverse crawling linked by dorsal and ventral turns.
  \textbf{c.}~The standard SLDS assumes Markovian transitions, which
  imply a geometric distribution of state durations (dotted black
  lines), but Fig.~\ref{fig:syllables} suggests that state
  transitions occur near the origin.  With the exception of state 7,
  which is unstable, the recurrent SLDS incorporates
  this dependency and generates state durations (solid black lines)
  that are a better fit to those inferred from the data (colored
  bars).  }
\label{fig:recurrent}
\end{figure}

@@

\begin{figure}[t!]
\centering
\includegraphics[width=6in]{figures/v3/figure7} 
\caption{\textit{Incorporating exogenous factors like oxygen level
    into the dynamics model yield differential responses in worms from
    different genetic strains and developmental stages.}
  \textbf{a.}~The observed neurons variances for each of the 41 worms
  in the~\citet{nichols2017global} dataset.  \textbf{b.}~Worms of
  different genetic strains (N2, npr1) and developmental stages
  (pre-lethargus, lethargus) exhibit different average state usages as
  a function of the oxygen level.  Blue and red states are much more
  likely in the lethargus stage for both genetic strains, but they are
  suppressed in npr1 worms at high oxygen concentrations. Moreover, we
  see a pronounced increase in the pink state usage immediately after
  the oxygen level increases to 21\%.  \textbf{c.}~As above, the
  inferred states closely match the manually labeled states
  of~\citet{nichols2017global}, with blue, red, and to a lesser extent
  green mapping onto the forward (\textsf{Fwd}) and quiescent
  (\textsf{Q}) states.  \textbf{d.}~We extended the model to
  incorporate oxygen level (O$_2$) and change in oxygen level
  ($\Delta$O$_2$) as exogenous factors that drive discrete transition
  probabilities.  We find that oxygen level has a limited effect on N2
  worms but strongly suppresses the blue (\textsf{FWD/Q}) state in
  npr1 worms and promotes the purple and orange
  (\textsf{Rev/DT/VT/Other}) states. The largest effect arises from
  the change in oxygen: an increase in $O_2$ strongly elicits the pink
  state.  \textbf{e.} The state-triggered average latent state
  trajectory following entry into the pink state shows a dip in
  activity along the last dimension 7 seconds after entry (black
  arrows).  \textbf{f.} This dip gives rise to strong opposing
  responses of sensory neurons \textsf{URXL/R} and \textsf{BAGL}
  (among others), which are known to respond to increases and
  decreases, respectively, in oxygen concentration.  }
\label{fig:o2}
\end{figure}

@@

>>>>

\begin{abstract}
  Recent advances in neural recording technologies have enabled
  simultaneous measurements of the majority of head ganglia neurons in
  both immobilized and freely-behaving
  \celegans~\citep{schrodel2013brain, prevedel2014simultaneous,
    nguyen2016whole}.  The dynamics of neural activity shed light on
  how \celegans~processes sensory information and generates motor
  activity.  To understand these dynamics, we develop recurrent state
  space models that decompose complex time-series into segments with
  simple, linear dynamics. We incorporate these models into a robust,
  hierarchical framework for combining information across whole-brain
  recordings of many worms.  Using this framework, we reveal latent
  states of population neural activity, along with the discrete
  behavioral ``syllables'' that drive dynamics in this latent state
  space.  We find stochastic transition patterns between these
  syllables, and we see that transition probabilities are determined
  by a combination of current brain state and environmental cues.  In
  addition to quantifying neural dynamics, this probabilistic
  framework aids in neural identification---currently a laborious,
  manual task---and reveals clusters of neurons that are similarly
  tuned in latent state space.  Finally, we find a significant overlap
  between our inferred syllables and the manually-labeled states
  of~\citet{kato2015global} and~\citet{nichols2017global}, which were
  shown to correspond to different behaviors, like forward crawling,
  reversals, and turns. Our methods automatically discover and
  quantify these behaviorally-meaningful states directly from neural
  activity, yielding powerful new tools for neuro-behavioral analysis.
\end{abstract}

@@

\textbf{Context:}
\begin{itemize}
\item Novel recording technologies enable simultaneous recording of
  O(100) of neurons simultaneously.  In \celegans, this constitutes
  the bulk of the head ganglia, and in other organisms this may constitue
  the majority of neurons in a specific circuit.
\item The promise is that these new recordings will lead to new
  insights into brain function. For example, with a complete view of a
  circuit, we should be able to study how it responds to external
  inputs, how it represents its internal state, and how it transforms
  internal states into circuit outputs.
\item \celegans~is the ideal model organism for realizing
  this goal.  It has a stereotyped nervous system with 302 identifiable
  neurons.  Its head ganglia is responsible for
  processing sensory inputs and generating outputs that initiate
  motor commands, like forward crawling, reversals, and turns.
\item With new recording capabilities, we can study how the
  coordinated activity of the \celegans~head ganglia encodes internal
  state, updates it in response to sensory information (like oxygen
  concentration), and transmits it via output (motor) neurons.
\item Moreover, we can study how the encoding, dynamics, and outputs
  differ across individual worms, genetic strains, and developmental stages.
\end{itemize}

\textbf{Three Problems: How to conceptualize the latent states of
  brain activity, their dynamics, and their relation to observed
  activity?  How to learn these dynamics from partial recordings of
  many individual worms? How to quantify differences between worms of
  different genetic strains and developmental stages?}
\begin{itemize}
\item We seek a model that is flexible enough to capture the complexity
  of neural dynamics and the relationship between latent states and
  observed activity, but we also require that the model be composed
  of interpretable building blocks. 
\item We have a spectrum of options --- simple factor analyzers and
  clustering algorithms, to Kalman filters and hidden Markov models,
  all the way up to highly nonlinear neural network models.
\item The challenge is to strike a balance: flexible yet interpretable.
\item While, in aggregate, we have a great deal of data to fit these models
  with, the recordings are spread over many individuals. In
  each individual we can only label a handful of neurons and we only
  record them for tens of minutes (photobleaching).
\item Moreover, each individual has its own idiosyncrasies due to
  factors like different levels of GCaMP expression from one worm
  to another, as well as subtle differences in the underlying brain
  dynamics. 
\item We need methods that combine these limited, partial recordings
  in order to learn shared structure shared across the population.
\item Finally, in order to understand the differences between
  genetic strains and developmental stages, we need models that allow
  us to flexibly test hypotheses about how they differ. 
\end{itemize}

\textbf{Main results, in broad brushstrokes:}
\begin{itemize}
\item We build a probabilistic framework for learning sophisticated neural dynamics from populations of worms. The framework is:
  \begin{itemize}
  \item \emph{hierarchical}, capturing canonical structure while allowing variability across worms, genetic strains, and developmental stages;
  \item \emph{robust}, ameliorating some concerns of model misspecification;
  \item \emph{recurrent}, capturing nuances in the co-evolution of discrete and continuous latent states;
  \item \emph{input-driven}, capturing the influence of exogenous factors and sensory inputs;
  \item and most importantly, \emph{composed} of simple and interpretable pieces that shed light on the dynamics of neural activity.
  \end{itemize}
\item We use this framework to extract low-dimensional state
  trajectories for each worm, using shared correlations across the
  population to align the trajectories and make predictions about
  unseen activity.
\item We discover clusters of similarity-tuned and highly correlated
  neurons, and we find that these neurons are functionally grouped
  according the different behaviors they are implicated in.
\item We show how our model's predictions can aid in identifying neurons (which will then drive down posterior uncertainty in the underlying dynamics of neural activity).
\item We show that brain dynamics are well-approximated as switching between discrete regimes, or ``syllables,'' each associated with simple linear dynamics.  These syllables map onto the manual segmentations of experimental experts, which were based on known behavioral implications of certain patterns of brain activity.
\item We quantify these linear dynamical regimes, show how they drive activity in clusters of neurons, and evaluate their differences from worm to worm.
\item We find that transitions between syllables are not simply Markovian; rather, the transitions depend on a combination of syllable history and latent brain state.
\item We measure how oxygen concentration (and change in oxygen concentration) differentially impact state transitions in worms of different genetic strains and developmental stages, and we find that sudden increases in oxygen elicit a characteristic, transient response regardless of genetic make-up or developmental stage, but that only certain genetic strains show a sustained response to increased oxygen levels. 
\end{itemize}

@@

We seek a parsimonious characterization of the dynamics of neural
activity in the head ganglia of \celegans. Recent advances in optical
imaging offer simultaneous recordings of most of the head ganglia
neurons, but we must surmount three main challenges in order to learn
a dynamics model from these data.  First, we need to design a model
flexible enough to capture nonlinear dynamics yet interpretable enough
to reveal simplifying structure when it exists.  Moreover, any model
will necessarily be an approximation to the true neural dynamics, and
as such we need to be robust to data that does not quite match model
predictions.  Finally, our model must combine partial recordings from
separate organisms to learn shared neural dynamics yet still allow for
individual variability. We construct a recurrent, robust, and
hierarchical state space model that addresses these challenges,
decomposing complex dynamics into a collection of simpler pieces that
can be learned from noisy, partial recordings with trial-to-trial
variability.  We call this model \textbf{TBD}.\todo{name it!}

We analyze the data presented in~\citet{kato2015global} and
\citet{nichols2017global}, which consist of a collection of optical
recordings of calcium fluorescence in immobilized \celegans. The
recordings are 18 minutes long with a frame rate of about 3Hz.  On
average, about 100 single units are extracted from each
recording. \celegans~is genetically stereotyped and each of its 302
neurons has a unique label (e.g. \textsf{AVAL}). On average, 30 single
units could be unambiguously labeled in each
recording. Fig.~\ref{fig:model}a illustrates this type of data. The
three panels represent recordings from three separate worms, and the
rows correspond to different neuron labels.  Neural activity is
measured for only a subset of neurons in each worm.

\begin{figure}[t!]
\centering%
\includegraphics[width=6in]{figures/v3/figure1} 
\caption{
  \textit{A probabilistic framework for combining
  information across multiple worms in order to learn a shared
  dynamics model for \celegans.}
  \textbf{a.}~The data consists of observed neural activity collected
  from many different worms.  In each recording we observe a different
  subset of labeled neurons in the head ganglia.
  \textbf{b.}~We model the observed activity as a linear function of
  an underlying, continuous, and typically low-dimensional latent state,
  here shown as a two-dimensional signal for each recording.
  \textbf{c.}~The dynamics of these continuous states are governed by
  a co-evolving discrete state.
  \textbf{d.}~These discrete states index into a library of linear
  dynamics functions for each worm, here shown as maps from~$\reals^2$ to~$\reals^2$.
  Each dynamics function is selectively used in different regions of
  continuous latent space, as shown by the different intensities of
  the vector field.
  \textbf{e.}~Though location in continuous state space helps inform
  transitions, the current discrete state does as well. This is captured
  by the Markov transition
  \textbf{e.}~To increase robustness to model misspecification, we
  use a multivariate-t distribution for the noise in the continuous
  state updates.
  \textbf{f.}~Each worm has its own dynamics library, but they are
  constrained to be close to the shared, canonical library.  This
  combines information to learn shared structure while also allowing
  for worm-to-worm variability.
  \textbf{g.} Likewise, we allow each worm to have its own observation
  variances to account for individual differences that may arise from,
  for example, variable levels of fluorescent protein expression.
  Missing neurons are denoted by gray boxes. 
}
\label{fig:model}
\end{figure}

@@

\begin{figure}[t!]
\centering
\includegraphics[width=6in]{figures/v4/figure2} 
\caption{ \textit{Hierarchical state space models smooth noisy
    recordings and predict the activity of unseen neurons.}  We show
  the neural activity (differenced Ca++ traces) for two of the
  five worms from the \citet{kato2015global} dataset.  Of these five
  worms, worm 1 (left) has the fewest labeled neurons and worm 5
  (right) has the most.  Green lines are the measured activity of the
  labeled neurons; black lines are the smoothed activities predicted
  by the model for the 61 neurons that were labeled in at least one
  worm.  The model uses correlations learned across all recordings to
  predict the activity of unlabeled neurons, like \textsf{SMBDL} in
  worm 1.  The neurons have been clustered based on similarity, as
  discussed in Fig.~\ref{fig:clustering}.  }
\label{fig:smoothing}
\end{figure}

@@

The degree to which the smoothed activity matches the observed
activity is governed by the dimensionality of the latent state space.
We estimate this dimensionality with cross validation, reserving the
last 20\% of the data (3.6 min) for testing and measuring the marginal
log likelihood of the held-out data as we increase the latent
dimensionality. We find that the test likelihood is maximized with a
10-dimensional latent state, and that variances differ substantially
between worms.

% Moreover, the likelihood is substantially higher in the model that
% includes per-worm observation variances compared to the model with
% shared variances for all worms. Fig.~\ref{fig:supp_emissions}d shows
% the inferred variance for each worm in the \citet{kato2015global}
% dataset.  Neurons that were not identified are marked with a gray box.
% This shows the extent of the missing data and the need for
% hierarchical modeling.\todo{make supplement figure}

@@

\begin{figure}[t!]
\centering
\includegraphics[width=6in]{figures/v4/figure3} 
\caption{ \textit{The mapping from continuous latent states to neural
    activity reveals clusters of functionally similar neurons.}
  \textbf{a.} The emission matrix is a linear map from latent states;
  the rows can be seen as ``tuning vectors'' that specify how each
  neuron responds to changes in the latent state.  These tuning
  vectors cluster into discrete groups (separated by black lines),
  indicating that there are subpopulations of similarly tuned, and
  hence highly correlated, neurons. Neurons are ordered as in
  Fig.~\ref{fig:smoothing}. \textbf{b.}~The pairwise similarity between
  tuning vectors supports this clustering and shows differences
  between subpopulations (clusters have been sorted to make this as
  close to diagonal as possible).  \textbf{c.}~The tuning similarity
  reflects correlations between neurons, but since it is learned as
  part of a hierarchical model, it predicts similarities for all pairs
  of neurons, even pairs that are never simultaneously recorded (gray,
  off-diagonal boxes).  \textbf{d.}~We show the spatial locations of
  neurons from each cluster, as specified by WormAtlas.  Neurons in
  the same cluster are not necessarily close to one another in space,
  though some clusters are spatially localized. Black bars denote
  approximate locations of the anterior (AG), dorsal (DG), ventral
  (VG) and retrovesicular ganglia.}
\label{fig:clustering}
\end{figure}

@@

\begin{figure}[t!]
\centering
\includegraphics[width=6in]{figures/v4/figure_4} 
\caption{ \textit{The predicted neural activity is a useful aid for
    neural identification.}  \textbf{a.} The model uses observed and
  labeled activity (green traces) to predict the activity of unlabeled
  neurons, as in Fig.~\ref{fig:smoothing}. \textbf{b.} Alongside the
  labeled activity, we also have traces from many neurons that could
  not be unambiguously labeled. To test whether the model's
  predictions could aid in labeling neurons, we conducted a synthetic
  experiment, withholding the activity of some labeled neurons. Here,
  the red trace is the withheld activity of VB01 and the blue traces
  are the activity of other unlabeled neurons.  Together, these make
  up a pool of candidates for the VB01 label.  \textbf{c.} For each
  candidate, we measured the correlation between its activity and the
  predicted activity of VB01.  \textbf{d.} Repeating this experiment
  50 times, we found that the correct candidates tend to be much more
  correlated with the model's predicted activity than the other,
  incorrect candidates.  Thus, the model's predictions serve as a
  useful aid for labeling neurons.  }
\label{fig:id}
\end{figure}

@@

\begin{figure}[t!]
\centering
\includegraphics[width=6in]{figures/v4/figure5} 
\caption{ \textit{The hierarchical SLDS automatically segments neural
    activity based on its dynamics.}  \textbf{a.}~The top three
  dimensions, according to variance explained, of the inferred
  continuous latent states for each of the five worms, color coded by
  the corresponding discrete latent state.  \textbf{b.}~The same
  continuous latent states but now color coded according to the manual
  segmentation of~\citet{kato2015global}.  \textbf{c.}~The inferred
  segmentations (top) and the manual segmentations (bottom) over time
  for worm 5. The segmentations largely align, though some manually
  labeled states are split into multiple inferred states (e.g. forest
  green state is often split into green and brown) and others are
  combined (e.g. light purple, light blue, and white are often
  combined into pink).  \textbf{d.}~We quantified the overlap between
  the manual and inferred states and found a close correspondence.
  As~\citet{kato2015global} showed, the manually labeled states map
  onto different behaviors (\textsf{REVSUS}: sustained reversal;
  \textsf{SLOW}: slow forward crawl; \textsf{VT}: ventral turn;
  \textsf{FWD}: forward crawl; \textsf{DT}: dorsal turn; \textsf{REV1,
    REV2}: reversals; \textsf{NOSTATE}: undetermined). The
  correspondence between inferred and manually labeled states implies
  that hierarchical state space models can infer behaviorally
  meaningful states directly from neural activity.}
\label{fig:syllables}
\end{figure}

@@

\begin{figure}[h!]
  \centering
  \vspace{-.5in}
  \includegraphics[width=6in]{figures/v4/figure6}
  \caption{\textit{Discrete states are characterized by
      simple linear dynamics that give rise to interpretable patterns
      of neural activity.}  \textbf{a.}~Each discrete state is
    associated with a canonical set of linear dynamics, which can be
    viewed as mappings from~$\reals^{10}$ to~$\reals^{10}$.  We show
    the mappings for the first three dimensions here, with vectors
    placed at locations where each state was used.  \textbf{b.}~We
    simulated four latent state trajectories with these linear
    dynamics from different starting locations, and then we mapped
    these latent state trajectories through the emission matrix to
    compute a time course of resulting activations for each cluster of
    neurons.}
  \label{fig:dynamics}
\end{figure}

@@

\begin{figure}[t!]
\centering
\includegraphics[width=6in]{figures/v4/figure7} 
\caption{ \textit{Discrete state transitions show a chain-like structure
    with non-Markovian durations.}  \textbf{a.}~Counts of inferred transitions
  from one state to another. The states are permuted to emphasize the
  feed-forward structure.  \textbf{b.}~An alternative view of the
  transitions; each state is a node in a graph and edges denote
  high-probability transitions.  This shows transitions between
  forward and reverse crawling linked by dorsal and ventral turns.
  \textbf{c.}~The standard SLDS assumes Markovian transitions, which
  imply a geometric distribution of state durations (dotted black
  lines), but Fig.~\ref{fig:syllables} suggests that state
  transitions occur near the origin.  With the exception of state 7,
  which is unstable, the recurrent SLDS incorporates
  this dependency and generates state durations (solid black lines)
  that are a better fit to those inferred from the data (colored
  bars).  }
\label{fig:recurrent}
\end{figure}

@@

\begin{figure}[t!]
\centering
\includegraphics[width=6in]{figures/v4/figure8} 
\caption{\textit{Incorporating exogenous factors like oxygen level
    into the dynamics model yield differential responses in worms from
    different genetic strains and developmental stages.}
  \textbf{a.}~The observed neurons variances for each of the 41 worms
  in the~\citet{nichols2017global} dataset.  \textbf{b.}~Worms of
  different genetic strains (N2, npr1) and developmental stages
  (pre-lethargus, lethargus) exhibit different average state usages as
  a function of the oxygen level.  Blue and red states are much more
  likely in the lethargus stage for both genetic strains, but they are
  suppressed in npr1 worms at high oxygen concentrations. Moreover, we
  see a pronounced increase in the pink state usage immediately after
  the oxygen level increases to 21\%.  \textbf{c.}~As above, the
  inferred states closely match the manually labeled states
  of~\citet{nichols2017global}, with blue, red, and to a lesser extent
  green mapping onto the forward (\textsf{Fwd}) and quiescent
  (\textsf{Q}) states.  \textbf{d.}~We extended the model to
  incorporate oxygen level (O$_2$) and change in oxygen level
  ($\Delta$O$_2$) as exogenous factors that drive discrete transition
  probabilities.  We find that oxygen level has a limited effect on N2
  worms but strongly suppresses the blue (\textsf{FWD/Q}) state in
  npr1 worms and promotes the purple and orange
  (\textsf{Rev/DT/VT/Other}) states. The largest effect arises from
  the change in oxygen: an increase in $O_2$ strongly elicits the pink
  state.  \textbf{e.} The state-triggered average latent state
  trajectory following entry into the pink state shows a dip in
  activity along the last dimension 7 seconds after entry (black
  arrows).  \textbf{f.} This dip gives rise to strong opposing
  responses of sensory neurons \textsf{URXL/R} and \textsf{BAGL}
  (among others), which are known to respond to increases and
  decreases, respectively, in oxygen concentration.  }
\label{fig:o2}
\end{figure}

@@

\section{Supplementary Figures}

\begin{itemize}
\item Graphical model (with arrows $y_t \to y_{t+1}$)
\item Neural activity for worms 2, 3, 4?
\item Model selection based on held-out log likelihood
\item Table of matching results?
\item Eigenspectra of inferred states
\item Model fitting 
\end{itemize}

\begin{figure}[t!]
\centering%
\includegraphics[width=5in]{figures/v3/figure1_supp} 
\caption{\textit{Graphical depiction of a hierarchical, robust, recurrent,
    and input-driven switching linear dynamical system (SLDS).}
  The blue arrow denotes the \textbf{hierarchical} component: each worm~$w$ has its
  own set of parameters~$\theta^{(w)}$, which contain the dynamics libraries, emission
  matrix, etc. These are slight perturbations of the same global mean parameters~$\theta$.
  In some cases, the worm-specific parameters are required to equal their global
  counterparts (as with the emission matrix). 
  Red arrows highlight the \textbf{robust} noise model that links~$x_t$ to~$x_{t+1}$.
  Green arrows indicate the \textbf{recurrent} extension: the next discrete syllable depends
  on both the preceding syllable as well as the preceding continuous latent state.
  Purple arrows carry the \textbf{input-dependence}: the next discrete syllable
  can also be governed by an exogenous input, like the oxygen level or the change
  in oxygen level. The remaining black arrows denote standard dependencies
  of the SLDS: discrete syllables and continuous latent states co-evolve, with
  the syllables specifying the linear dynamics to use when updating the continuous
  latent states.  The observations are a function of these underlying latent states.
  The only non-standard addition is an autoregressive dependency between~$y_t$ and
  $y_{t+1}$; here, the SLDS models only the differences in observed fluoresence
  value between two consecutive frames.
}
\label{fig:graphical_model}
\end{figure}

\begin{figure}[t!]
\centering%
\includegraphics[width=6in]{figures/v4/figure5_supp1} 
\caption{\textit{Model selection on the basis of marginal log
    likelihood of test data.}  We withheld the last 20\% of the
  continuous latent states from each worm in order to compare
  different models.  For each combination of hierarchical
  vs. standard, robust vs standard, and recurrent vs. standard, we
  swept from~$K=2$ syllables to~$K=20$ (from white to purple)
  syllables in steps of size two. We also included the special case
  of~$K=1$ (grey bars), which reduces to standard autoregressive
  models. For each combination, we fit the model and evaluated the
  marginal likelihood of the held-out test data. The best-performing
  model was the hierarchical, robust, recurrent SLDS with $K=8$
  syllables.}
\label{fig:model_selection}
\end{figure}

\begin{figure}[t!]
\centering%
\includegraphics[width=6in]{figures/v4/figure5_supp2} 
\caption{\textit{Discrete states carry substantial information about
    continuous latent states.} How much information is carried by the
  discrete syllables alone? In blue we show the mean and two standard deviations of
  the predictive distribution~$p(x \given z)$.  The esimated values of~$x$
  are shown in black. Below we show the inferred syllables~$z$. We see that
  the discrete syllable carry a surprising amount of information about the continuous
  latent states.}
\label{fig:z_predictions}
\end{figure}
